Functory: Une bibliothèque de calcul distribué \\ pour Objective Caml\footnotetext{Ce travail a été réalisé en partie dans le cadre du projet U3CAT (\emph{Unification of Critical C Code Analysis Techniques}, ANR-08-SEGI-021).}

### 2010-10-22 doc/jfla-2011/main.tex
*** empty log message ***

\hrulefill ICI\hrulefill

Les deux modules \of{Cores} et \of{Network}

\begin{flushleft}
  \quad  \textbf{tant que} tâches à faire $\lor$ tâches en cours \\
  \quad  \quad \textbf{tant que} tâches à faire $\land$ ouvriers disponibles \\
  \quad  \quad \quad affecter une tâche à un ouvrier \\
  \quad  \quad \textbf{attendre} la fin d'une tâche \\
  \quad  \quad \quad ajouter les nouvelles tâches renvoyées par \of{master} \\
\end{flushleft}

@@

The \of{Cores} module implements the distributed computing library for
multiple cores on the same machine. 
% As illustrated in Section~\ref{sec:example}, 
It provides a function
\of{set_number_of_cores: int -> unit} to indicate the number of cores
to be used. The number passed to this function may be different from
the actual number of cores in the machine; it is rather the number of
tasks to be performed simultaneously.
% TODO comment more on the number of cores?
% Thus a program using the \of{Cores} module typically begins with 
% \begin{ocaml}
% open Cores
% let () = set_number_of_cores 8
% \end{ocaml}

The \of{Cores} module is implemented with \unix\ processes, using the
\of{fork} and \of{wait} system calls provided by the \of{Unix} library
of \Ocaml. The idea is pretty simple.  The \of{compute} function
maintains a global table of pending tasks and keeps track of the
number of idle cores.  Whenever there is a pending task and an idle
core, a new sub-process is created using \of{Unix.fork}; once
completed, the sub-process marshals the result into a 
local file. The
main process maintains a table mapping each sub-process ID to the
input task and the local file name. It waits for any completed
sub-process using \of{Unix.wait} and recovers the result from the
local file. Then function \of{master} is applied, which may generate
new tasks.  The main loop can be depicted in the following way:

We also have an alternative implementation using
\unix\ pipes instead of local files.

The scheduling of tasks to the different cores is left to the
operating system, through \of{Unix.fork}. Thus it may be the case that
two tasks are executed on the same core, even if the declared number
of cores is less or equal than the actual number of cores on the machine.
% TODO: comment on user-provided scheduling?

@@

The \of{Network} module implements the distributed computing library
for a network of machines. 
% As illustrated in Section~\ref{sec:example}, 
It provides a function
\of{declare_workers: n:int -> string -> unit} to fill a table of
worker machines. 
% The argument \of{n} indicates the number of cores to
% be used in each worker machine, analogous to the \of{Cores}
% module. Thus a program using the \of{Network} module typically begins
% with
% \begin{ocaml}
% open Network
% let () = declare_workers ~n:12 "hostname1"
% let () = declare_workers ~n:8  "192.168.24.2"
% ...
% \end{ocaml}

The \of{Network} module is based on a traditional TCP-based client/server
architecture, where each worker is a server and the master is the
client of each worker. The main execution loop is similar to the one
in the \of{Cores} module, where distant processes on remote machines
correspond to sub-processes and idle cores are the idle cores 
of remote workers. In addition to this, it also involves issues of message
transfer and fault tolerance, which are subsequently described.

\subsubsection{Protocol}\label{sec:protocol}

Messages sent from master to workers could be any of the following kinds:
\begin{description}
\item[\of{Assign(id:int, f:string, x:string)}] This message assigns a
  new task to the worker, the task being identified by the unique
  integer \of{id}. The task to be performed is given by strings \of{f}
  and \of{x}, which are interpreted depending on the context.

\item[\of{Kill(id:int)}] This message tells the worker to kill the task
  identified by \of{id}.

\item[\of{Stop}] This message informs the worker about completion of
  the computation, so that it may choose to exit.

\item[\of{Ping}] This message is used to check if the worker is still
  alive, expecting a \of{Pong} message from the worker in return.
\end{description}
Messages sent by workers could be any of the following kinds:
\begin{description}
\item[\of{Pong}] This message is an acknowledgment for a \of{Ping}
  message from the master.

\item[\of{Completed(id:int, s:string)}] This message indicates the
  completion of a task identified by \of{id}, with result \of{s}.

\item[\of{Aborted(id:int)}] This message informs the master that the
  task identified by \of{id} is aborted, either as a response to a
  \of{Kill} message or because of a worker malfunction.
\end{description}
Our implementation of the protocol works across different
architectures, so that master and workers could be run on completely
different platforms w.r.t. endianness, version of \Ocaml\ and
operating system.

\subsubsection{Network Sub-modules} % TODO: change title

As mentioned in Section~\ref{sec:API}, the \of{Network} module
actually provides three different implementations, according to three
different execution scenarios. There are provided in three
sub-modules, as described below.

\paragraph{\of{Same}.} This module is used when master and workers are
running the same executable. The master and workers have to be
differentiated in some manner. We use an environment variable
\texttt{WORKER} for this purpose. When set, it indicates that the
executable acts as a worker. At run-time, a worker immediately 
enters a loop waiting for tasks from the master, without even getting
into the user code. 
As explained in Section~\ref{sec:API}, 
the master function has the following signature.
\begin{ocaml}
 val compute : 
      worker:('a -> 'b) -> 
      master:('a * 'c -> 'b -> ('a * 'c) list) -> ('a * 'c) list -> unit  
\end{ocaml}
The master uses marshaling to send both a closure of type 
\of{'a -> 'b} and a task of type \of{'a} to the worker. The resulting strings
are passed as argument \of{f} and \of{x} in message
\of{Assign}. Similarly, the worker uses marshaling to send back the
result of the computation of type \of{'b}, which is the argument
\of{s} in message \of{Completed}.

Though the ability to run the same executable helps a lot in deploying
the program in different machines, it comes at a small price. Since
the worker is not getting into the user code, closures which are
transmitted from the master cannot refer to global variables in the
user code. Indeed, the initialization code for these global variables
is never reached on the worker side. For instance, the Mandelbrot set
example could be written as follows:
\begin{ocaml}
  let max_iterations = 200
  let worker (xmi, xma, ymi, yma, w, h) =
    ... draw sub-image using max_iterations ...
\end{ocaml}
That is, the global function \of{worker} makes use of the global
variable \of{max_iterations}. The worker gets the function to compute
from the master, namely the closure corresponding to function
\of{worker} in that case, but on the worker side the initialization of
\of{max_iterations} is never executed.

One obvious solution is  not to use global variables in the
worker code. This is not always possible, though. 
To overcome this, the \of{Same} sub-module also provides a
\of{Worker.compute} function to start the worker loop manually from
the user code. This way, it can be started at any point, in particular
after the initialization of the required global variables.
Master and worker are still running the same executable, but are 
distinguished using a user-defined way (command-line argument,
environment variable, etc.).

\bigskip
There are situations where it is not possible
to run the same executable for master and workers.
For instance, architectures or operating systems could be different
across the network.
For that reason, the \of{Network} module provides two other
implementations.

\paragraph{\of{Poly}.} When master and workers are compiled with the
same version of \Ocaml, we can no longer marshal closures but we can
still marshal polymorphic values. Indeed, an interesting property of
marshaling in \Ocaml\ is to be fully architecture-independent, as long
as a single version of \Ocaml\ is used. It is worth pointing out that
absence of marshaled closures
now enables the use of two different programs for master and workers. This
is not mandatory, though, since master and workers could still be
distinguished at run-time as in the previous case.

On the worker side, the main loop is started manually using
\of{Worker.compute}. The computation to be performed on each task is
given as an argument to this function. It thus looks as follows:
\begin{ocaml}
  Worker.compute : ('a -> 'b) -> unit -> unit
\end{ocaml}
On the master side, the \of{compute} function is simpler than in the
previous case, as it has one argument less, and thus has the following
signature. 
\begin{ocaml}
  Master.compute : master:('a * 'c -> 'b -> ('a * 'c) list) -> ('a * 'c) list -> unit 
\end{ocaml}
For realistic applications, where master and workers are completely
different programs, possibly written by different teams, this is the
module of choice in our library, since it can still pass polymorphic
values over the network. The issues of marshaling are automatically
taken care of by \Ocaml\ run-time.

The derived API presented in Section~\ref{sec:derived} is adapted to
deal with the absence of closures. Exactly as the \of{compute}
function, each API now takes two forms, one for the master and another
for the workers. For example, \of{map_fold_ac} takes the following
forms.
\begin{ocaml}
  Worker.map_fold_ac : map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> unit -> unit
  Master.map_fold_ac : 'b -> 'a list -> 'b
\end{ocaml}
It is the responsibility of the user to ensure type consistency between
master and workers.

\paragraph{\of{Mono}.} When master and workers are compiled using
different versions of \Ocaml, we can no longer use marshaling. 
As in the previous case, we split \of{compute} into two functions, one
for master and one for workers. In addition, values transmitted over
the network can only be strings. The signature thus takes the
following form.
\begin{ocaml}
  Worker.compute : (string -> string) -> unit -> unit
  Master.compute : master:(string * 'c -> string -> (string * 'c) list) -> 
                  (string * 'c) list -> unit
\end{ocaml}
Any other datatype for tasks should be encoded to/from strings. This
conversion is left to the user.
Note that the second component of each task is still polymorphic (of
type \of{'c} here), since it is local to the master.

% There is
% actually a single, generic implementation, which is then instantiated
% in three different ways, corresponding to the three scenarios.  The
% generic implementation assumes that only strings are being transmitted
% over the network, as evident from the \of{Assign} and \of{Completed}
% messages in Section~\ref{sec:protocol}. Depending on the
% implementation, these strings could be marshaled closures, marshaled
% values or simply user strings.

\subsubsection{Fault Tolerance}\label{sec:fault}

@@

% LocalWords:  sérialisation boutisme inatteignables observationnellement
% LocalWords:  sous-modules

>>>>

Les deux modules \of{Cores} et \of{Network} contiennent une boucle
principale analogue, de la forme suivante :
\begin{flushleft}
  \quad  \textbf{tant que} tâches à faire $\lor$ tâches en cours \\
  \quad  \quad \textbf{tant que} tâches à faire $\land$ ouvriers disponibles \\
  \quad  \quad \quad affecter une tâche à un ouvrier \\
  \quad  \quad \textbf{attendre} la fin d'une tâche \\
  \quad  \quad \quad ajouter les nouvelles tâches renvoyées par \of{master} \\
\end{flushleft}
La différence se situe dans la technologie utilisée pour affecter une
tâche à un ouvrier, pour attendre la fin d'une tâche et enfin pour
ajouter de la tolérance aux pannes.

@@

Le module \of{Cores} permet de distribuer le calcul sur une unique
machine, typiquement en exploitant plusieurs \coeurs. 
Comme illustré section~\ref{sec:studies}, la
\of{set_number_of_cores} permet de spécifier le nombre de \coeurs\ à
utiliser. Ce nombre peut être différent du nombre effectif de \coeurs\ 
de la machine. Il peut même être strictement plus grand.
En fait, ce nombre indique tout simplement combien de tâches peuvent être
effectuées simultanément.

Le module \of{Cores} est réalisé à l'aide de processus \unix, en
utilisant la fonction \of{Unix.fork} de la bibliothèque standard d'\Ocaml.
Plus précisément, un compteur indique le nombre d'ouvriers disponibles 
et l'affectation d'une tâche à un ouvrier se fait en 
créant un nouveau sous-processus avec \of{fork}.
À la fin du calcul, le résultat de la tâche est transmis au processus
père par sérialisation dans un tube.
La répartition des tâches sur les différents \coeurs\ (physiques) de
la machine, le cas échéant, est laissée au système. Il se peut donc
que deux tâches se retrouvent être exécutées sur le même \coeur, y
compris dans le cas où le nombre de \coeurs\ déclarés est inférieur ou
égal au nombre de \coeurs\ effectifs.

@@

Le module \of{Network} permet de distribuer le calcul sur un réseau de
machines. 
Comme illustré section~\ref{sec:studies}, la fonction 
\of{declare_workers: n:int -> string -> unit} permet de déclarer
l'ensemble des machines du réseau sur lesquels se trouvent des
ouvriers, en spécifiant un nombre d'ouvriers par machine (qui 
ne coïncide pas nécessairement avec un nombre effectif de \coeurs).
Le module \of{Network} est basé sur une architecture client/serveur
utilisant TCP/IP, où chaque ouvrier est un serveur et où le patron
est (ironiquement) le client de chaque ouvrier.

\paragraph{Protocole.}\label{sec:protocol}
Le protocole utilisé comporte sept messages différents.
Les quatre messages possibles envoyés par le patron à un ouvrier sont : 
\of{Assign(id:int, f:string, x:string)} pour affecter une nouvelle
tâche \of{id} à l'ouvrier, sous la forme de deux chaînes \of{f} et \of{x} dont
le sens dépend du contexte ; 
\of{Kill(id:int)} pour demander l'interruption de la tâche \of{id} ;
\of{Stop} pour demander l'arrêt définitif de l'ouvrier ;
et enfin \of{Ping} pour vérifier que l'ouvrier est toujours réactif.
Inversement, les trois messages possibles envoyés par un ouvrier au
patron sont :
\of{Pong} en réponse à un message \of{Ping} ;
\of{Completed(id:int, s:string)} pour renvoyer le résultat \of{s} de
la tâche \of{id} ;
et enfin \of{Aborted(id:int)} pour signifier l'échec de la tâche
\of{id}, soit en réponse à \of{Kill}, soit à cause d'une erreur lors
de l'exécution de la fonction \of{worker}.

Ce protocole est de telle
sorte que patron et ouvriers peuvent être exécutés sur des
plate-formes complètement différentes, vis-à-vis de
l'\emph{endianness}, de la version d'\Ocaml\ et du système d'exploitation.
Dans le sous-module \of{Same}, les arguments \of{f} et \of{x} du
message \of{Assign} désignent respectivement
la sérialisation d'une fonction et de
son argument, et l'argument \of{s} du message \of{Completed} la
sérialisation du résultat. Dans les sous-module \of{Poly} et
\of{Mono}, en revanche, la
partie \of{f} du message \of{Assign} n'est plus pertinente
car la fonction \of{worker} est maintenant locale à l'ouvrier.
Les chaînes \of{x} et \of{s} restent pertinentes ; pour \of{Poly},
ce sont des sérialisations de valeurs \Ocaml\ et pour \of{Mono}  de
simples chaînes de caractères.

\subsubsection{Tolérance aux pannes}\label{sec:fault}

@@

% LocalWords:  sérialisation boutisme inatteignables observationnellement
% LocalWords:  sous-modules sérialisations
